---
date: "4/20/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)
library(broom)
library(mosaic)
library(infer)

#read insurance data file
insurance_data = read.csv("~/Downloads/insurance.csv") 

#shorten name
insurance = insurance_data

#examine data
head(insurance)
dim(insurance)
names(insurance)

#create histogram of charges to understand overall distribution
ggplot(insurance, aes(x = charges)) +
  geom_histogram() + 
  labs(x = "Amount Charged", y = "Frequency")

# The distribution is clearly heavily skewed to the right, and is unimodal
# with what appears to be a small number of upper outliers
# to remove some of this skew, we will take the logarithm of charges

#mutate the data to take logarithms
insurance = insurance %>%
  mutate(logcharges = log(charges))

ggplot(insurance, aes(x = logcharges)) +
  geom_histogram() + 
  labs(x = "Log Amount Charged", y = "Frequency")



# One can now see that the data looks far more normal and easy to work with

# Create boxplots for charges v categorical variables (sex, smoker)

ggplot(insurance, aes(y = logcharges, x = sex)) + 
  geom_boxplot()

# Sex does not seem to impact insurance charges, although the variance for 
# males is slightly larger

ggplot(insurance, aes(y = logcharges, x = smoker)) +
  geom_boxplot()

# There is a clear relationship that smokers incur larger insurance charges

ggplot(insurance, aes(y = logcharges, x = children)) +
  geom_boxplot()

# Create scatterplots of charges v age, charges v bmi, charges v children

ggplot(insurance, aes(y = charges, x = age)) +
  geom_point(alpha = 0.4)

# There appears to be an extremely consistent linear relationship between
# insurance charges and age, possibly because older individuals are at 
# higher risk of incurring medical cost. 

ggplot(insurance, aes(y = charges, x = bmi))+
  geom_point(alpha = 0.4)

# Interestingly, there appears to be little overall correlation between
# charges and BMI, although there do seem to be 2 distinct groups
# one with a positive linear relationship and one with no relationship

ggplot(insurance, aes(y = charges, x = bmi, color = smoker))+
  geom_point(alpha = 0.4)

# Controlling for smoking, one sees that although there is nor relationship
# between charges and BMI for nonsmokers, there is a positive correlation between
# charges and BMI for smokers. There is also a jump at bmi = 30 for smokers
# that is not there for nonsmokers, that for the most part seems to double
# the overall cost

ggplot(insurance, aes(y = logcharges, x = children)) +
  geom_point(alpha = 0.4)

# Although variance increases as number of children decreases, there does not
# appear to be a linear correlation between charges and number of children

```

Now that the data has been analyzed and visualized, we will create models based
off of the approximately linear relationships viewed. This includes charges
v BMI for smokers and charges v age. So that the models may be better compared 
and combined, we will only use data form smokers for the charges v age model as 
well.

```{r}
# Subset the data to be smokers only
insurance_smokers = insurance %>%
  filter(smoker == "yes")

# Create model of bmi v charges and print summary
bmi_model = lm(charges ~ bmi, data = insurance_smokers)
summary(bmi_model)

# Assumption 1 is met because it is reasonable to assume one person's bmi and 
# insurance charge is independent from another's

# Draw the regression line
ggplot(insurance_smokers, aes(y = charges, x = bmi)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1473.11, intercept = -13186, color = "blue")

# Create a data frame with actual values, predicted values, and residuals from 
# model

bmi_mod_results = data.frame(observed = insurance_smokers$charges,
                             predicted = bmi_model$fitted.values,
                             residual = bmi_model$residuals)

# Plot the residuals versus the predictions
ggplot(bmi_mod_results, aes(y = residual, x = predicted)) +
  geom_point() + 
  geom_hline(yintercept = 0)

# The residual plot does not look great overall. There are clear trends, with the
# residuals decreasing from predcicted = 9.5 - 10.25, a jump at 10.25, and then
# decreasing again from 10.5 - 11.5. As such, this model isn't great for the 
# data, due to the jump at bmi = 30


# Create a Q-Q plot of the residuals
ggplot(bmi_mod_results, aes(sample = residual)) +
  geom_qq()

# Despite the failings of the residual plot, the Q-Q plot actually look pretty
# good, and is an approximately straight line

```

Next, we segment the data into bmi > 30 and bmi < 30 to try to remove some of
the failings of the previous model, using the same steps

```{r}

# Filter the data by bmi greater than and less then or equal to 30

lowbmi_smokers = insurance_smokers %>%
  filter(bmi <= 30)

highbmi_smokers = insurance_smokers %>%
  filter(bmi > 30)

# Follow the same steps for model creation and assumption checking

lowbmi_model = lm(charges ~ bmi, data = lowbmi_smokers)
summary(lowbmi_model)

highbmi_model = lm(charges ~ bmi, data = highbmi_smokers)
summary(highbmi_model)

ggplot(lowbmi_smokers, aes(y = charges, x = bmi)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 485.1, intercept = 9071.2, color = "blue")

ggplot(highbmi_smokers, aes(y = charges, x = bmi)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 508.5, intercept = 23617.2, color = "blue")


lowbmi_mod_results = data.frame(observed = lowbmi_smokers$charges,
                             predicted = lowbmi_model$fitted.values,
                             residual = lowbmi_model$residuals)

highbmi_mod_results = data.frame(observed = highbmi_smokers$charges,
                             predicted = highbmi_model$fitted.values,
                             residual = highbmi_model$residuals)

ggplot(lowbmi_mod_results, aes(y = residual, x = predicted)) +
  geom_point() + 
  geom_hline(yintercept = 0)

ggplot(highbmi_mod_results, aes(y = residual, x = predicted)) +
  geom_point() + 
  geom_hline(yintercept = 0)

# Both of these residual plots look significantly better, with the points
# symmetrically clustered around low values of y with no clear pattern.
# However, there is still failings to the residual plots, as in both, there are
# numerous data points with extremely high residuals, but a lack of the same 
# outliers on the lower end

ggplot(lowbmi_mod_results, aes(sample = residual)) +
  geom_qq()

ggplot(highbmi_mod_results, aes(sample = residual)) +
  geom_qq()

# The Q-Q plot  looks worse for the more segmented models, perhaps
# because the variance in the overall data tends to grow far greater as 
# bmi approaches 30

# From this analysis, it is clear that neither the segmented nor the
# unsegmented (by bmi) data conforms perfectly to a linear model. Quality
# of the models will be assessed later in the project

```

Next, we create the charges v age model and assess the assumptions

```{r}
# Follow the same steps for model creation and assumption checking

age_model = lm(charges ~ age, data = insurance_smokers)
summary(age_model)

# Assumption 1 is met because it is reasonable to assume one person's age and 
# insurance charge is independent from another's

ggplot(insurance_smokers, aes(y = charges, x = age)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 305.24, intercept = 20294, color = "blue")

# It is interesting to note that there are two distinct groups in the data,
# each of which seems to have a nearly identical linear relationship.
# None of the other variables in the dataset can account for these two groups,
# so their cause is unknown, but it most likely has something to do with risk.
# We sould expect the residual plot to be symmetric and randomly distriuted,
# but focused towards values of y farther from 0, as the model produces an 
# estimate that is in the middle of the two distinct groups.

age_mod_results = data.frame(observed = insurance_smokers$charges,
                             predicted = age_model$fitted.values,
                             residual = age_model$residuals)

# Plot the residuals versus the predictions
ggplot(age_mod_results, aes(y = residual, x = predicted)) +
  geom_point() + 
  geom_hline(yintercept = 0)

# The residual plot looks exactly as expected and described in the previous
# block comment, with residual values far from 0, but otherwise randomly
# and symmetrically distributed


# Create a Q-Q plot of the residuals
ggplot(age_mod_results, aes(sample = residual)) +
  geom_qq()

# The Q-Q plot consists of 2 distinct sections, one for negative residual
# values and one for positive, each of which looks like a moderate straight
# line (and therefore fulfills assuption of 2). To make the residual and 
# Q-Q plots perfect, one must subset the data by the two distinct groups in the 
# scatterplot, which cannot be done given that there is no obvious reason
# for these 2 groups in the dataset. Because of the issues in the residual
# and Q-Q plots, neither of these models perfectly fulfill assumption 2.

```

Next, we create a multivariate linear model using both age and bmi

```{r}

# Follow the same steps for model creation and assumption checking

combined_model = lm(charges ~ age + bmi, data = insurance_smokers)
summary(combined_model)


combined_mod_results = data.frame(observed = insurance_smokers$charges,
                             predicted = combined_model$fitted.values,
                             residual = combined_model$residuals)

ggplot(combined_mod_results, aes(y = residual, x = predicted)) +
  geom_point() + 
  geom_hline(yintercept = 0)

# The residual plot has values symmetrically distributed about 0, with low y 
# values, however there is in obvious pattern, with a break in the middle
# most likely caused by the break in the age. However, of all the residual
# plots, this fits the requirements the closest


# Create a Q-Q plot of the residuals
ggplot(combined_mod_results, aes(sample = residual)) +
  geom_qq()

# The Q-Q plot for the model of combined bmi and age looks correct, with 
# a single straight line, signifying that the combined model meets assumption
# 2 better than any of the other previous models
```

To conclude our work with these models, we assess the quality of each model created using MSPE and R-squared.The R-squared values are  taken from the summaries printed above

```{r}
# bmi model

#MSPE
mean(bmi_mod_results$residual^2)
"R^2: 0.6504"

# lowbmi model

# MSPE
mean(lowbmi_mod_results$residual^2)
"R^2: 0.09283"

# highbmi model

# MSPE
mean(highbmi_mod_results$residual^2)
"R^2: 0.1333" 

# age model

# MSPE
mean(age_mod_results$residual^2)
"R^2: 0.1356"

# combined model

# MSPE
mean(combined_mod_results$residual^2)
"R^2: 0.7532"

# From this information, especially that of the R^2 values, combined
# with information from the assumptions section, we conclude that the 
# combined model is by far the best model for linear prediction because it has
# the highest R^2 value, and fits the assumptions the closest (although not 
# perfectly)
```

To conclude the project, we will use a 90% confidence interval to predict
the average cost of insurance for a nonsmoker older than 40 with a bmi under 30.

```{r}
# Create a dataset of nonsmokers older than 40 with a bmi under 30

confidence_data = insurance %>%
  filter(smoker == "no") %>%
  filter(bmi < 30) %>%
  filter(age >= 40) 

dim(confidence_data)

# Set a random number seed

set.seed(2021)

# Take a sample of 150 people from the dataset

sample = sample(confidence_data, size = 150, replace = FALSE)

head(sample)

# Calculate the sample mean charge

sample_mean = mean(sample$charges)
sample_sd = sd(sample$charges)

print(sample_mean)
print(sample_sd)

# n > 40, so we are able to calculate the confidence interval for the mean 
# charge using the sample mean and standard deviation, using the z value for 
# a 90% CI, 1.645, and the formula:
# x_bar ±1.645(s/sqrt(n))

lower_bound = sample_mean - 1.645 * (sample_sd/sqrt(150))
upper_bound = sample_mean + 1.645 * (sample_sd/sqrt(150))

print(lower_bound)
print(upper_bound)

# The 90% confidence interval for expected charge for a nonsmoker with bmi
# under 30 and age over 40 is [10751.93, 12120.89]






